Safe Transferable Regions

### 2016-07-01 paper/introduction.tex
fixed typos minor corrections

\paragraph{Safe Transferable Regions.}
In this paper, we present a language-based approach to safely and
efficiently realize such transferable data that uses regions as
the implementation technique.
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

>>>>

\paragraph{Safe Transferable Regions.}
In this paper, we present a language-based approach to cleanly encapsulated
transferable data and a safe and efficient implementation of such data based
on regions. 
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

### 2016-07-01 paper/type-inference.tex
fixed typos minor corrections

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an \C{A} object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the constructor and method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as p@$\rgn_0$ in
    letregion $\rgn_1$ in
      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}

>>>>

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair} object \C{q}, and an object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as p@$\rgn_0$ in
    letregion $\rgn_1$ in
      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}

### 2016-07-01 paper/type-system.tex
fixed typos minor corrections

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
\FB, hence it has no region annotations.
% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
defines a type rule for such values, allowing them to be typed.
Type safety theorem is now stated thus:
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}

>>>>

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
\FB, hence it has no region annotations.
% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
% \vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
defines a type rule for such values, allowing them to be typed.
The following type safety theorem shows that a well-typed program will never
attempt to dereference an ``invalid'' reference (a reference to an object
in a region that has been transferred or freed):
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}
